Untwisting the Boerdijk--Coxeter Helix

### 2017-04-19 doc/TetrahelixUntwist.tex
lame-o attempt to add a wobble diagram

\title{Untwisting the Boerdijk--Coxeter}
\author{Robert L. Read \email{read.robert@gmail.com}
}

@@

\Cref{fig:helixcomparison} illustrates this difference. Although the nodes conincide, \cref{graycoxeter} evaluated
at real values generates the black helix, and \cref{eq:colored} defines the red, yellow, and blue helices. (In this figure
these rail helices have been rendered at a slightly higher radius than the nodes; in actuality the helices are much close
to the edges between the nodes that they nearly match.)

@@

The $H_{BCcolored}(n,c)$ formulation can be further clarified by rewriting directly in terms of the rail angle $\rho_{bc}$ rather than $\theta_{bc}$.
Intuitively we seek an expression where $c/3$ is multiplied by a $1/3$ rotation plus the rail angle $\rho$.
We expand 
the expressions $\theta_{bc}$ and $\rho_{bc}$ in \eqref{eq:colored} and seek to isolate the term $c2\pi/3 $.
\begin{align*}
 c \theta_{bc}  &=   \text{\{we aim for 3 in denominator, so we split...\}} \\
    (c/3)  (3 \theta_{bc})  &=   \text{\{we want $2\pi$ in numerator, so add canceling terms...\}} \\
 (c/ 3) ((3 \theta_{bc} - 2 \pi)  + 2 \pi) &= \text{\{definition of $\rho_{bc}$\}...} \\  
  (c / 3) \rho_{bc}  + c 2 \pi /3 &=  \text{algebra...} \\  
c  ( \rho_{bc} +  2 \pi) /3  \\
\end{align*}

@@

The \emph{pitch} of any tetrahelix, defined as the length of a complete revolution
where $\rho \neq 0$ is:
\begin{equation}
  \label{pitcheqn}
p(\rho) = \frac{2 \pi  \cdot d}{\rho}
\end{equation}

@@

We use the term \emph{tetrahelix} to mean any structure made of
vertices and edges which is isomorphic to the BC helix and in which the
vertices lie on three helices. We further demand that all edge lengths
be finite, as we are only interested in physically constructable tetrahelices.
By isomporphic we mean there is a one-to-one mapping between both
vertices and edges in the two tetrahelices.
One could consider various definitions of optimality for a
tetrahelix, but the most useful to us as robotocists working with the Tetrobot
concept is to minimize the
maximum difference between any two edges, because the Tetrobot uses
mechanical linear acutators with
limited range of extension.

A \emph{triple helix} is three congruent helices that share an axis. We show that
optimal tetrahelices are triple helices.

Suppose that all three rails do not have the same pitch. Starting at any shortest edge, as we move
from node to node away from our start node edge lengths between rails must always lengthen without bound,
which cannot be optimal.
So we are justified in talking about the
\emph{pitch} of 
the optimal tetrahelix as the pitch of its three rail helices, even though there are
three such helices of equivalent pitch.

@@

Since the axes are parallel, we may define the \emph{inradius}, represented by the letter $i$, of a
tetrahelix to be the radius of the largest
cylinder parallel to this axis contained within the circumscribing cylinder
which is penetrated by no edge.
Define a \emph{minimax edge-length optimal tetrahelix} or just an
\emph{optimal tetrahelix} to be a tetrahelix for which there exists
no other tetrahelix of the same inradius and pitch with a lower maximum
difference between its edge lengths.

@@

Since all three rails have the same rail length, no matter how we
move the rails in the $xy$ plane if we shorten the $xy$ distance between
vertices we shorten the total distance.
Our tool for thinking about this is to collapse the $z$ dimension to form
a two-dimensional figure of nodes and non-rail edges (see \cref{untouchingrailfig}.)
Consider the projection along the $z$ axis of all vertices and non-rail edges into the $xy$ plane, which will be
a figure of dots and connecting segments in the $xy$ plane. The convex
hull for any one helix projection will be a circle (if its pitch is
irrational) or a polygon if rational, or a point if the helix has
zero curvature. Each of these figures by definition lies outside the
circle of inradius $r_{in}$ in the $xy$ plane.

Case 1: Suppose that $\rho$ is zero. Then for any given inradius, an equilateral
triangle is the minimax solution for all non-rail edges. Since there is only
one rail-edge length, this is the minimax solution for the entire set. Since
the vertices of an equilateral triangle lies on a circle, all points lie on
a cylinder.

Case 2: Suppose that $\rho$ is positive but less than $\pi$.
In this case each rail helix has
curavature and places points on both sides of any line through the origin
of the $xy$ plane (or both coincident on such a line.)

Supoose there is a rail $P$ which does not have a segment touching the incircle
at all in the projection, as depicted in \cref{untouchingrailfig}.
Then a segment connecting the other two $Q$ and $R$ is a longest
rail, because any chord touching an incircle is longer than any which does not.
These two rails ($Q$ and $R$) can be moved closer to each other, and further away from $P$, to form a better minimax solution
by shortening the longest rail, until one of the segments from $P$ the previously
untouching rail touches one of the incircle. By induction, then all rails have
at least one segement touching each vertex on the rail that is tangent to the
incircle in an optimal tetrahelix.

Now suppose that you attempt to move the axis of one of the rails. Because
$\rho > 0 $ and $\rho$ is not a multiple of $\pi$, there is some vertex on the two sides of any diameter
halving our circle. Moving the axis of a tetrahelix lengthens a longest line while ``pinching'' the
inradius on the other side, so it is not optimal. Since the axes of these helices are the
same and they have the same curvature and pitch, the points defined by the helices all lie
on a cylinder.

@@

@@

\begin{theorem}
  \label{eventhirds}
  An optimal tetrahelix of any rail angle $\rho < \pi$ is a triple helix all nodes evenly spaced at $d/3$ intervals on the $z$ axis.
  Any one tetrahedron in a tetrahelix has $1$ rail edge, $2$ one-hop edges connected to the rail and $2$ two-hop edges connected to the rail.
  The edge opposite of the rail edge is a one-hop edge.
\end{theorem}

@@

\begin{theorem}
  \label{generalformulaoptimal}
  The tetrahelices generated by $H_{general}$ are optimal in terms of minimum maximum member length when $r_{\rho}$ is chosen so that
  the length of the one-hop edge is equal to the rail length.
\end{theorem}

@@

If we look down the axis of an optimal tetrahelix as shown in \cref{axialview}, it happens that only one of the one-hop edges
(rendered in purple in our software)
comes closest to the axis. In other words, they define the radius of the incircle of the
projection, or the radius of a cylinder that would just fit inside the tetrahelix.
A formula for the inradious of the tetrahelix is useful if you are designing it as a structure that bears something internally,
such as a firehose, a pipe, or
a ladder for a human. The inradius $r_{in}(\rho)$ of
an optimal tetrahelix is a remarkably simple function of the radius $r$ and the rail angle $\rho$:
\begin{equation}
  \label{eq:inradius}
  r_{in}(\rho) = r \sin{\frac{\pi - \rho}{6}}
\end{equation}
Which can be seen from the trigonometry of a diagram of the projected one-hop edges
connecting four sequentially numbered vertices:

@@

\begin{corollary}
  The equitetrabeam with minimal maximal edge difference is produced
  \newline by $H_{general}$ when $ r = \sqrt{\frac{8}{27}} $.
\end{corollary}

@@

The curvature of a rail helix is given by:
\begin{equation}
  \frac{|r_{\rho}|}{r_{\rho}^2 + (d_{\rho}/\rho)^2}
\end{equation}
which goes to $0$ as $\rho$ approaches $0$ (the equitetrabeam.)
As $\rho$ increase up to $\rho_{bc}$ the curvature increases smoothly until the BC Helix is reached.

@@

If an equitetrabeam were actually
used as a beam, an engineer might start with the optimal tetrabeam and
dilate it in one dimension to ``deepen'' the beam. Similarly, simple
length changes curve the equitetrabeam into an ``arch''.
The ``colored'' approach of \cref{eq:general} exposes these possibilities
more than the approach of \cref{eq:coxetergraygeneral}.

@@

The BC Helix is the end point of a continuum of tetrahelices, the
other end point being an untwisted tetrahelix with equilateral cross
section, constructed by changing the length of only those members
crossing the outside rails after hopping over the nearest
vertex. Under the condition of miminum maximum length difference of
all members in the system, all such tetrahelices have vertices evenly
spaced along the axis generated by a simple equation.  A mechanical
machine, such as a robot or a variable-geometry truss, that can change
the length of its members, can thus twist and untwist itself by
changing the length of the appropriate members to achieve any point in
the continuum optimally. With a numeric solution, a design may choose
a rotation angle and member lengths to obtain any desired pitch.

@@

@@

Improve the helix comparison by constructing by hand in the Inkscape.

Check the math on V_{general}.

@@

>>>>

\title{Untwisting the Boerdijk--Coxeter Helix}
\author{Robert L. Read \email{read.robert@gmail.com}
}

@@

\Cref{fig:helixcomparison} illustrates this difference with a 7-tetrahedra BC helix, which is
in fact the same geometry as the robot illustrated in \cref{fig:glussbot}.
Although the nodes conincide,
\cref{graycoxeter} evaluated
at real values generates the black helix which runs through every node, and \cref{eq:colored} defines
the red, yellow, and blue helices. (In this figure
these rail helices have been rendered at a slightly higher radius than the nodes for clarity;
in actuality the maximum distance between the continuous, curved helix and the
straight edges between nodes is much smaller than can be clearly rendered.)

@@

The $H_{BCcolored}(n,c)$ formulation can be further clarified by rewriting directly in terms of the rail angle $\rho_{bc}$ rather than $\theta_{bc}$.
Intuitively we seek an expression where $c/3$ is multiplied by a $1/3$ rotation plus the rail angle $\rho$.
We expand 
the expressions $\theta_{bc}$ and $\rho_{bc}$ in \eqref{eq:colored} and seek to isolate the term $c2\pi/3 $.
\begin{align*}
 c \theta_{bc}  &=   \text{\{we aim for 3 in denominator, so we split...\}} \\
    (c/3)  (3 \theta_{bc})  &=   \text{\{we want $2\pi$ in numerator, so add canceling terms...\}} \\
 (c/ 3) ((3 \theta_{bc} - 2 \pi)  + 2 \pi) &= \text{\{definition of $\rho_{bc}$\}...} \\  
  (c / 3) \rho_{bc}  + c 2 \pi /3 &=  \text{\{algebra...\}} \\  
c  ( \rho_{bc} +  2 \pi) /3  \\
\end{align*}

@@

The \emph{pitch} of any tetrahelix, defined as the axial length of a complete revolution
where $\rho \neq 0$ is:
\begin{equation}
  \label{pitcheqn}
p(\rho) = \frac{2 \pi  \cdot d}{\rho}
\end{equation}

@@

We use the term \emph{tetrahelix} to mean any structure made of
vertices and edges which is isomorphic to the BC helix and in which the
vertices lie on three helices. We further demand that all edge lengths
be finite, as we are only interested in physically constructable tetrahelices.
By isomporphic we mean there is a one-to-one mapping between both
vertices and edges in the two tetrahelices.
One could consider various definitions of optimality for a
tetrahelix, but the most useful to us as robotocists working with the Tetrobot
concept is to minimize the
maximum ratio between any two edge lengths, because the Tetrobot uses
mechanical linear acutators with
limited range of extension.

A \emph{triple helix} is three congruent helices that share an axis. We show that
optimal tetrahelices are in fact triple helices with the same radius, so that all
vertices are on a cylinder. In, stages, we demonstrate that optimal tetrahelices:
\begin{enumerate}
\item have the same pitch,
\item have parallel axes,
\item share the same axis,
\item have the same radius.
\end{enumerate}

Suppose that all three rails do not have the same pitch. Starting at any shortest edge between
two rails, as we move
from node to node away from our start edge the edge lengths
between rails must always lengthen without bound,
which cannot be optimal.
So we are justified in talking about the
\emph{pitch} of 
the optimal tetrahelix as the pitch of its three rail helices, even though there are
three such helices of equivalent pitch.

@@

Since the axes are parallel, we may define the \emph{inradius}, represented by the letter $i$, of a
tetrahelix to be the radius of the largest
cylinder parallel to this axis that is surrounded by each tetrahelix and pentrated by no edge.
Define a \emph{minimax edge-length optimal tetrahelix} or just an
\emph{optimal tetrahelix} to be a tetrahelix for which there exists
no other tetrahelix of the same inradius and pitch with a lower maximum
ratio between its edge lengths.

@@

Since all three rails have the same rail edge length, no matter how we
move the rails in the $xy$ plane if we shorten the $xy$ distance between
vertices we shorten the total distance.
Our tool for thinking about this is to collapse the $z$ dimension to form
a two-dimensional figure of nodes and non-rail edges (see \cref{untouchingrailfig}.)
Consider the projection along the $z$ axis of all vertices and non-rail edges into the $xy$ plane, which will be
a figure of dots and connecting segments in the $xy$ plane. The convex
hull for any one helix projection will be a circle (if its pitch is
irrational) or a polygon if rational, or a point if the helix has
zero curvature. Each of these figures by definition lies outside the
circle of inradius $r_{in}$ in the $xy$ plane.

As a little lemma for the proof below, observe that a tetrahelix of zero radius, where
all points lie on the same line,
is not as optimal as a tetrahelix of a small radius. The edges between rails will be
shorter than the rail edges, and moving them apart slightly lengthens the between-edge
rails.

\begin{theorem}
  Any optimal tetrahelix with a rail angle of magnitude less than $\pi$ has all three
  axes conincident.
  \label{thm:conincident}
\end{theorem}
\begin{proof}
  Case 1: Suppose that $\rho$ is zero. Then for any given inradius, the
  figure in the $XY$-plane of an equilateral
  triangle is the minimax solution for all non-rail edges. Since all
  rail edges are of length 1, this is the minimax solution for the entire set. Since
the vertices of an equilateral triangle lie on a circle, all points in three-space lie on
a cylinder.

Case 2: Suppose that $\rho$ is positive but less than $\pi$.
In this case each rail helix has
curavature. The projection of points in the $XY$ plane creates a figure
guaranteed to have point on either side of any line through the axis of such
a helix, because the figure is either an $n$-gon or a circle.
We show that the three helices share a common axis.

  Without loss of generality define the Red helix to have its axis on the $z$-axis.
  Without loss of generatlity define the Blue helix to be a helix that
  has an edge connection to the Red helix that is either a maximum or a minimum.
  Let $B'$ be the blue helix translated in the $XY$-plane so that its axis is the $z$-axis and
  condicent with the red helix $R$. Let $D$ be the distance between the axis of the Blue helix
  $B$ and $B'$. We will show that if $D>0$ then $B$ ``wobbles'' in a way that cannot be optimal.
  Define a wobble vector by:
  \begin{equation*}
    W(n) = B(n) - B'(n)
  \end{equation*}
  where $B(n)$ is the cartesian vector
  $\begin{bmatrix}
           x \\
           y 
  \end{bmatrix}$
  for the projection of the $n$th blue vertex.
  Note that $|| R(n) - B'(n+k)||$ is a constant for any $k$, because $R$ and $B'$ have the
  same pitch and the same axis, even if they do not have the same radius
  (which we are currently proving.)

  \begin{figure}[H]
     \centering
     \includegraphics[width=0.7\textwidth]{figures/WobbleDiagram.png}
     \caption{Wobble Vectors from Non-Coincident Axes}
  \label{fig:wobble}
\end{figure}

  Since $-\pi<\rho<\pi$, each rail helix rotates about its axis as a function of $n$ and
  defines at least 3 unique points when projected onto the $XY$-plane around its axis.
  If $2\pi/\rho$ is irrational, it defines a circle. If $2\pi/\rho$ is rational it defines
  an n-gon in the $XY$-plane, where $n$ is at least 3 (when $\rho = 2\pi/3$).
  The set of wobbles $\{W(n) | \text{for any n}\}$ thus contains at least three vectors,
  pointing in different directions.
  For any point not at the origin, at least one of these vectors moves closer to the
  point and at least one moves further away.

    The set of all lengths in the tetrahelix is a superset of:
    $L = \{|| R(n) - B(n)||\}$, which by our choice has at least one longest or shortest
    length.
    $L = \{|| R(n) - (B'(n) + W(n))||\}$ and so
    $L = \{|| (R(n) - B'(n)) - W(n)||\}$.
    But $R(n) - B'(n)$ is a constant, so the minimax value of $L$ is improved as $||W(n)||$
    decreases.  
    By our choice, this improves the minimax value of the total tetrahelix.

    This process can be carried out on both the Blue and Yellow helices
    (perhaps simultaneously) until $W(n)$ is
    zero for both, finding a tetrahelix of improved overall minimax value at each step.
    So a tetrahelix is optimal only when $W(n) = 0$, and therefore when $D=0$
    $B(n) = B'(n)$, all three axes are conincident.
  \end{proof}

@@

Note: This is not a proof that they have the same radius, I will have to add that.

@@

\begin{theorem}
  \label{eventhirds}
  An optimal tetrahelix of any rail angle $\rho < \pi$ is a triple helix with all vertices evenly spaced at $d/3$ intervals on the $z$ axis.
  Any one tetrahedron in a tetrahelix has $1$ rail edge, $2$ one-hop edges connected to the rail and $2$ two-hop edges connected to the rail.
  The edge opposite of the rail edge is a one-hop edge.
\end{theorem}

@@

\begin{theorem}
  \label{generalformulaoptimal}
  The tetrahelices generated by $H_{general}$ are optimal in terms of minimum maximum ratio of member length when $r_{\rho}$ is chosen so that
  the length of the one-hop edge is equal to the rail length.
\end{theorem}

@@

If we look down the axis of an optimal tetrahelix as shown in \cref{axialview}, it happens that only one of the one-hop edges
(rendered in purple in our software)
comes closest to the axis. In other words, they define the radius of the incircle of the
projection, or the radius of a cylinder that would just fit inside the tetrahelix.
A formula for the inradius of the tetrahelix is useful if you are designing it as a structure that bears something internally,
such as a firehose, a pipe, or
a ladder for a human. The inradius $r_{in}(\rho)$ of
an optimal tetrahelix is a remarkably simple function of the radius $r$ and the rail angle $\rho$:
\begin{equation}
  \label{eq:inradius}
  r_{in}(\rho) = r \sin{\frac{\pi - \rho}{6}}
\end{equation}
Which can be seen from the trigonometry of a diagram of the projected one-hop edges
connecting four sequentially numbered vertices:

@@

\begin{corollary}
  The equitetrabeam with minimal maximal edge ratio is produced
  \newline by $H_{general}$ when $ r = \sqrt{\frac{8}{27}} $.
\end{corollary}

@@

The curvature of a rail helix is foramlly given by:
\begin{equation}
  \frac{|r_{\rho}|}{r_{\rho}^2 + (d_{\rho}/\rho)^2}
\end{equation}
which goes to $0$ as $\rho$ approaches $0$ (the equitetrabeam.)
As $\rho$ increase up to $\rho_{bc}$ the curvature increases smoothly until the BC Helix is reached.

@@

If an equitetrabeam were actually
used as a beam, an engineer might start with the optimal tetrabeam and
dilate it in one dimension to ``deepen'' the beam. Similarly, simple
length changes curve the equitetrabeam into an ``arch''.
The ``colored'' approach of \cref{eq:general} exposes these possibilities
more than the approach of \cref{graycoxeter}.

@@

The BC Helix is the end point of a continuum of tetrahelices, the
other end point being an untwisted tetrahelix with equilateral cross
section, constructed by changing the length of only those members
crossing the outside rails after hopping over the nearest
vertex. Under the condition of miminum maximum length ratios of
all members in the system, all such tetrahelices have vertices evenly
spaced along the axis generated by a simple equation and are in fact triple helices.
A machine, such as a robot or a variable-geometry truss, that can change
the length of its members can thus twist and untwist itself by
changing the length of the appropriate members to achieve any point in
the continuum. With a numeric solution, a design may choose
a rotation angle and member lengths to obtain a desired pitch.

@@

** Try to show that rotation must be symmetric over possible rotation.

** Try to show 2-hop > rail for any optimal.

@@

@@

Case 1: Suppose that $\rho$ is zero. Then for any given inradius, an equilateral
triangle is the minimax solution for all non-rail edges. Since there is only
one rail-edge length, this is the minimax solution for the entire set. Since
the vertices of an equilateral triangle lie on a circle, all points lie on
a cylinder.

Case 2: Suppose that $\rho$ is positive but less than $\pi$.
In this case each rail helix has
curavature and places points on both sides of any line through its axis.

Supoose there is a rail $P$ which does not have a segment touching the incircle
at all in the projection, as depicted in \cref{untouchingrailfig}.
Then a segment connecting the other two $Q$ and $R$ is a longest
rail, because any chord touching an incircle is longer than any which does not.
These two rails ($Q$ and $R$) can be moved closer to each other, and further away from $P$,
to form a better minimax solution
by shortening the longest rail, until one of the segments from $P$, the previously
untouching rail, touches the incircle. By induction, then all rails have
at least one segement touching each vertex on the rail that is tangent to the
incircle in an optimal tetrahelix.

Now suppose that you attempt to move the axis of one of the rails away from the center
for our circle. Because
$\rho > 0 $ and $\rho$ is not a multiple of $\pi$, there is some vertex on the two sides of any diameter
halving our circle. Moving the axis of a tetrahelix lengthens a longest line while ``pinching'' the
inradius on the other side, so it is not optimal. Since the axes of these helices are the
same and they have the same curvature and pitch, the points defined by the helices all lie
on a cylinder.
\end{proof}
